Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2022-03-12 paper/paper.tex
update

\begin{example}\label{eg:twonode}
In the following, we show a minimum example of mapping the independent problem of a 2 vertex graph (left) to a tensor network (right).

@@

\section{Maximum independent sets and its counting}\label{sec:counting}
\subsection{Tropical algebra for finding the MIS size and counting MISs}
In the previous section, we focused on computing the independence polynomial for a graph $G$ of a given MIS size $\alpha(G)$, but we did not show how to compute this number.
The method we use to compute this quantity is based on the following observations. Let $x=\infty$, the independence polynomial becomes
\begin{equation}
I(G, \infty) = a_{\alpha(G)} \infty^{\alpha(G)},
\end{equation}
where the lower-order terms vanish. We can thus replace the polynomial type $a = (a_0, a_1, \ldots, a_k)$ with a new type with two fields: the largest exponent $k$ and its coefficient $a_k$.
From this, we can define a new algebra as
\begin{equation}
    \eqname{P1}
\begin{split}
    a_x\infty^x \oplus a_y\infty^y &= \begin{cases}
        (a_x + a_y)\infty^{\max(x,y)}, & x = y\\
        a_y\infty^{\max(x,y)}, & x < y\\
        a_x\infty^{\max(x,y)}, & x > y
    \end{cases}, \\
    a_x\infty^x \odot a_y\infty^y &= a_x a_y\infty^{x+y}\\
    \mymathbb{0} &= 0\infty^{-\infty}\\
    \mymathbb{1} &= 1\infty^{0}.
\end{split}%& \text{\rlap{(P1)}}
\label{eq:countingtropical}
\end{equation}
To implement this algebra programmatically, we create a data type with two fields $(x, a_x)$ to store the MIS size and its counting,
and define the above operations and constants correspondingly.
If one is only interested in finding the MIS size, one can drop the counting field.
The algebra of the exponents becomes the max-plus tropical algebra~\cite{Maclagan2015, Moore2011}:
\begin{equation}\eqname{T}
    \begin{split}
        x \oplus y &= \max(x,y)\\
        x \odot y &= x + y\\
        \mymathbb{0} &= -\infty\\
        \mymathbb{1} &= 0.
    \end{split}\label{eq:tropical}
\end{equation}
This algebra is the same as the one used in Liu et al.~\cite{Liu2021} to calculate and count spin glass ground states.
For independent set calculations here, the vertex tensor and edge tensor becomes:
\begin{equation}
    W^{\rm T} = \left(\begin{matrix}
        \mymathbb{1} \\
        \infty
    \end{matrix}\right),   
    \qquad \qquad
        B^{\rm T} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}

\subsection{Truncated polynomial algebra for counting independent sets of large size}
Instead of counting just the MISs, one may be interested in counting the independent sets of large sizes close to the MIS size.
For example, if one is interested in counting only $a_{\alpha(G)}$ and $a_{\alpha(G)-1}$, we can define a truncated polynomial algebra by keeping only the largest two coefficients in the polynomial in \Eq{eq:polynomial} as:
\begin{equation}
    \eqname{P2}
    \begin{split}
    a \oplus b &= (a_{\max(k_a, k_b)-1} + b_{\max(k_a, k_b)-1}, a_{\max(k_a, k_b)} + b_{\max(k_a, k_b)}),\\
    a \odot b &= (a_{k_a-1} b_{k_b}+a_{k_a} b_{k_b-1}, a_{k_a} b_{k_b}),\\
    \mymathbb{0} &= (), \\
    \mymathbb{1} &= (1).\label{eq:max2poly}
    \end{split}
\end{equation}
In the program, we thus need a data structure that contains three fields, the largest order $k$, and the coefficients for the two largest orders $a_k$ and $a_{k-1}$.
This approach can clearly be extended to calculate more independence polynomial coefficients and is more efficient than calculating the entire independence polynomial.
As will be shown below, this algebra can also be extended to enumerate those large-size independent sets.

>>>>

\begin{example}\label{eg:twonode}
In the following, we show a minimum example of mapping the independent problem of a 2 vertex complete graph K2 (left) to a tensor network (right).

@@

\section{Maximum independent sets and its counting}\label{sec:counting}
\subsection{Tropical algebra for finding the MIS size and counting MISs}
In the previous section, we focused on computing the independence polynomial for a graph $G$ of a given MIS size $\alpha(G)$, but we did not show how to compute this number.
The method we use to compute this quantity is based on the following observations. Let $x=\infty$, the independence polynomial becomes
\begin{equation}
I(G, \infty) = a_{\alpha(G)} \infty^{\alpha(G)},
\end{equation}
where the lower-order terms vanish. We can thus replace the polynomial type $a = (a_0, a_1, \ldots, a_k)$ with a new type with two fields: the largest exponent $k$ and its coefficient $a_k$.
From this, we can define a new algebra as
\begin{equation}
    \eqname{P1}
\begin{split}
    a_x\infty^x \oplus a_y\infty^y &= \begin{cases}
        (a_x + a_y)\infty^{\max(x,y)}, & x = y\\
        a_y\infty^{\max(x,y)}, & x < y\\
        a_x\infty^{\max(x,y)}, & x > y
    \end{cases}, \\
    a_x\infty^x \odot a_y\infty^y &= a_x a_y\infty^{x+y}\\
    \mymathbb{0} &= 0\infty^{-\infty}\\
    \mymathbb{1} &= 1\infty^{0}.
\end{split}%& \text{\rlap{(P1)}}
\label{eq:countingtropical}
\end{equation}
To implement this algebra programmatically, we create a data type with two fields $(x, a_x)$ to store the MIS size and its counting,
and define the above operations and constants correspondingly.
If one is only interested in finding the MIS size, one can drop the counting field.
The algebra of the exponents becomes the max-plus tropical algebra~\cite{Maclagan2015, Moore2011}:
\begin{equation}\eqname{T}
    \begin{split}
        x \oplus y &= \max(x,y)\\
        x \odot y &= x + y\\
        \mymathbb{0} &= -\infty\\
        \mymathbb{1} &= 0.
    \end{split}\label{eq:tropical}
\end{equation}
Algebra \Eq{eq:tropical} and \Eq{eq:countingtropical} are the same as those used in Liu et al.~\cite{Liu2021} to calculate and count spin glass ground states.
For independent set calculations here, the vertex tensor and edge tensor becomes:
\begin{equation}
    W^{\rm T} = \left(\begin{matrix}
        \mymathbb{1} \\
        \infty
    \end{matrix}\right),   
    \qquad \qquad
        B^{\rm T} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}

\subsection{Truncated polynomial algebra for counting independent sets of large size}
Instead of counting just the MISs, one may be interested in counting the independent sets of large sizes close to the MIS size.
For example, if one is interested in counting only $a_{\alpha(G)}$ and $a_{\alpha(G)-1}$, we can define a truncated polynomial algebra by keeping only the largest two coefficients in the polynomial in \Eq{eq:polynomial} as:
\begin{equation}
    \eqname{P2}
    \begin{split}
    a \oplus b &= (a_{\max(k_a, k_b)-1} + b_{\max(k_a, k_b)-1}, a_{\max(k_a, k_b)} + b_{\max(k_a, k_b)}),\\
    a \odot b &= (a_{k_a-1} b_{k_b}+a_{k_a} b_{k_b-1}, a_{k_a} b_{k_b}),\\
    \mymathbb{0} &= (), \\
    \mymathbb{1} &= (1).\label{eq:max2poly}
    \end{split}
\end{equation}
In the program, we thus need a data structure that contains three fields, the largest order $k$, and the coefficients for the two largest orders $a_k$ and $a_{k-1}$.
This approach can clearly be extended to calculate more independence polynomial coefficients and is more efficient than calculating the entire independence polynomial.
Similarly, one can also truncate the polynomial and keep only its smallest several orders, which can be used for e.g. counting the minimum maximal independent sets.
As will be shown below, this algebra can also be extended to enumerate those large-size independent sets.
